---
title: "notebook_1"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

# notebook_1.Rmd is inside /notebooks, so ".." means project root
PROJECT_ROOT <- normalizePath("..", winslash = "/", mustWork = TRUE)
setwd(PROJECT_ROOT)

source("src/00_project_setup.R")
source("src/01_packages.R")
source("src/02_helpers.R")

# Save notebook outputs without changing notebook display
NOTEBOOK_OUTPUT_DIR <- file.path(PROJECT_ROOT, "notebooks", "notebook_1_outputs")
NOTEBOOK_CSV_DIR <- file.path(NOTEBOOK_OUTPUT_DIR, "csv")
NOTEBOOK_PLOT_DIR <- file.path(NOTEBOOK_OUTPUT_DIR, "plots")

fs::dir_create(NOTEBOOK_OUTPUT_DIR)
fs::dir_create(NOTEBOOK_CSV_DIR)
fs::dir_create(NOTEBOOK_PLOT_DIR)
```

# Notebook 1: Data Understanding and Type Validation

This notebook checks whether the processed soccer datasets load correctly, have the expected column classes, contain plausible values, and obey the basic match-outcome logic needed for modeling.

This is an inspection notebook, so it is compiled as HTML instead of PDF. HTML handles wide tables much better.

All displayed validation tables are also saved under:

```text
notebooks/notebook_1_outputs/csv/
```

# 1. Processed files

```{r processed-files}
processed_files <- c(
  statsbomb_competitions = file.path(PROCESSED_DIR, "statsbomb_competitions.csv"),
  statsbomb_matches = file.path(PROCESSED_DIR, "statsbomb_matches.csv"),
  football_data_uk_matches = file.path(PROCESSED_DIR, "football_data_uk_matches.csv"),
  international_results = file.path(PROCESSED_DIR, "international_results.csv")
)

file_check <- tibble::tibble(
  dataset = names(processed_files),
  path = unname(processed_files),
  exists = file.exists(processed_files),
  file_size_mb = round(file.info(processed_files)$size / 1024^2, 3)
)

readr::write_csv(
  file_check,
  file.path(NOTEBOOK_CSV_DIR, "01_processed_files_check.csv")
)

file_check
```

```{r stop-if-missing-files}
if (any(!file_check$exists)) {
  stop("One or more processed files are missing. Run the full pipeline before knitting this notebook.")
}
```

# 2. Load processed data

Each file is loaded using `processed_csv_col_types()` from `src/02_helpers.R`.

```{r load-processed-data}
processed_data <- purrr::imap(
  processed_files,
  ~ readr::read_csv(
    file = .x,
    col_types = processed_csv_col_types(.x),
    show_col_types = FALSE
  )
)

loaded_data_names <- tibble::tibble(dataset = names(processed_data))

readr::write_csv(
  loaded_data_names,
  file.path(NOTEBOOK_CSV_DIR, "02_loaded_data_names.csv")
)

names(processed_data)
```

# 3. Dataset inventory

```{r dataset-inventory}
dataset_inventory <- purrr::imap_dfr(
  processed_data,
  function(dat, dataset_name) {
    tibble::tibble(
      dataset = dataset_name,
      rows = nrow(dat),
      columns = ncol(dat),
      first_date = if ("date" %in% names(dat)) as.character(min(dat$date, na.rm = TRUE)) else NA_character_,
      last_date = if ("date" %in% names(dat)) as.character(max(dat$date, na.rm = TRUE)) else NA_character_
    )
  }
)

readr::write_csv(
  dataset_inventory,
  file.path(NOTEBOOK_CSV_DIR, "03_dataset_inventory.csv")
)

dataset_inventory
```

# 4. All columns, classes, and examples

This table shows every column, its R class, storage type, missing count, unique count, and example values.

```{r class-summary}
class_summary <- purrr::imap_dfr(
  processed_data,
  function(dat, dataset_name) {
    tibble::tibble(
      dataset = dataset_name,
      column_position = seq_along(names(dat)),
      column = names(dat),
      class = purrr::map_chr(dat, ~ paste(class(.x), collapse = " / ")),
      typeof = purrr::map_chr(dat, typeof),
      n_missing = purrr::map_int(dat, ~ sum(is.na(.x))),
      n_unique = purrr::map_int(dat, dplyr::n_distinct),
      example_values = purrr::map_chr(
        dat,
        ~ paste(utils::head(unique(na.omit(.x)), 5), collapse = ", ")
      )
    )
  }
)

readr::write_csv(
  class_summary,
  file.path(NOTEBOOK_CSV_DIR, "04_class_summary.csv")
)

class_summary
```

# 5. Expected versus actual classes

```{r expected-classes}
expected_common_match_classes <- tibble::tribble(
  ~column, ~expected_class,
  "source", "character",
  "raw_file", "character",
  "source_match_id", "character",
  "date", "Date",
  "season", "character",
  "competition", "character",
  "home_team", "character",
  "away_team", "character",
  "home_score", "integer",
  "away_score", "integer",
  "match_result", "character",
  "result_class", "integer",
  "home_win", "integer",
  "draw", "integer",
  "away_win", "integer",
  "goal_difference", "integer",
  "total_goals", "integer",
  "neutral", "logical"
)

expected_classes <- dplyr::bind_rows(
  expected_common_match_classes |>
    dplyr::mutate(dataset = "statsbomb_matches"),

  expected_common_match_classes |>
    dplyr::mutate(dataset = "football_data_uk_matches"),

  expected_common_match_classes |>
    dplyr::mutate(dataset = "international_results"),

  tibble::tribble(
    ~column, ~expected_class,
    "competition_id", "integer",
    "season_id", "integer",
    "country_name", "character",
    "competition_name", "character",
    "competition_gender", "character",
    "competition_youth", "logical",
    "competition_international", "logical",
    "season_name", "character",
    "match_updated", "character",
    "match_updated_360", "character",
    "match_available_360", "character",
    "match_available", "character"
  ) |>
    dplyr::mutate(dataset = "statsbomb_competitions")
) |>
  dplyr::select(dataset, column, expected_class)

readr::write_csv(
  expected_classes,
  file.path(NOTEBOOK_CSV_DIR, "05_expected_classes.csv")
)

expected_classes
```

```{r class-validation}
class_validation <- class_summary |>
  dplyr::select(
    dataset,
    column,
    actual_class = class,
    typeof,
    n_missing,
    n_unique,
    example_values
  ) |>
  dplyr::left_join(
    expected_classes,
    by = c("dataset", "column")
  ) |>
  dplyr::mutate(
    expected_class = dplyr::coalesce(expected_class, "not specified"),
    class_ok = dplyr::case_when(
      expected_class == "not specified" ~ NA,
      actual_class == expected_class ~ TRUE,
      stringr::str_starts(actual_class, paste0(expected_class, " /")) ~ TRUE,
      TRUE ~ FALSE
    )
  ) |>
  dplyr::arrange(dataset, dplyr::desc(class_ok == FALSE), column)

readr::write_csv(
  class_validation,
  file.path(NOTEBOOK_CSV_DIR, "05_class_validation.csv")
)

class_validation
```

```{r failed-class-checks}
failed_class_checks <- class_validation |>
  dplyr::filter(class_ok == FALSE)

readr::write_csv(
  failed_class_checks,
  file.path(NOTEBOOK_CSV_DIR, "05_failed_class_checks.csv")
)

failed_class_checks
```

# 6. Compact examples from each dataset

These examples use selected columns so they are readable.

## 6.1 StatsBomb competitions

```{r examples-statsbomb-competitions}
examples_statsbomb_competitions <- processed_data$statsbomb_competitions |>
  dplyr::slice_head(n = 10)

readr::write_csv(
  examples_statsbomb_competitions,
  file.path(NOTEBOOK_CSV_DIR, "06_1_examples_statsbomb_competitions.csv")
)

examples_statsbomb_competitions
```

## 6.2 StatsBomb matches

```{r examples-statsbomb-matches}
examples_statsbomb_matches <- processed_data$statsbomb_matches |>
  dplyr::select(
    source_match_id,
    date,
    season,
    competition,
    home_team,
    away_team,
    home_score,
    away_score,
    match_result,
    result_class,
    neutral
  ) |>
  dplyr::slice_head(n = 10)

readr::write_csv(
  examples_statsbomb_matches,
  file.path(NOTEBOOK_CSV_DIR, "06_2_examples_statsbomb_matches.csv")
)

examples_statsbomb_matches
```

## 6.3 football-data.co.uk matches

```{r examples-football-data-uk}
examples_football_data_uk <- processed_data$football_data_uk_matches |>
  dplyr::select(
    source_match_id,
    date,
    season,
    competition,
    home_team,
    away_team,
    home_score,
    away_score,
    match_result,
    result_class,
    full_time_result,
    home_shots,
    away_shots,
    home_shots_on_target,
    away_shots_on_target
  ) |>
  dplyr::slice_head(n = 10)

readr::write_csv(
  examples_football_data_uk,
  file.path(NOTEBOOK_CSV_DIR, "06_3_examples_football_data_uk.csv")
)

examples_football_data_uk
```

## 6.4 International results

```{r examples-international-results}
examples_international_results <- processed_data$international_results |>
  dplyr::select(
    source_match_id,
    date,
    season,
    tournament,
    home_team,
    away_team,
    home_score,
    away_score,
    match_result,
    result_class,
    neutral,
    city,
    country
  ) |>
  dplyr::slice_head(n = 10)

readr::write_csv(
  examples_international_results,
  file.path(NOTEBOOK_CSV_DIR, "06_4_examples_international_results.csv")
)

examples_international_results
```

# 7. Random examples from match datasets

First rows can be misleading. Random examples help check whether values look plausible across the whole file.

```{r random-match-examples}
set.seed(2026)

match_tables <- processed_data[c(
  "statsbomb_matches",
  "football_data_uk_matches",
  "international_results"
)]

random_match_examples <- purrr::imap_dfr(
  match_tables,
  function(dat, dataset_name) {
    dat |>
      dplyr::select(
        source_match_id,
        date,
        competition,
        home_team,
        away_team,
        home_score,
        away_score,
        match_result,
        result_class,
        home_win,
        draw,
        away_win,
        goal_difference,
        total_goals,
        neutral
      ) |>
      dplyr::slice_sample(n = min(10, nrow(dat))) |>
      dplyr::mutate(dataset = dataset_name, .before = 1)
  }
)

readr::write_csv(
  random_match_examples,
  file.path(NOTEBOOK_CSV_DIR, "07_random_match_examples.csv")
)

random_match_examples
```

# 8. Outcome logic validation

This verifies that the derived outcome columns agree with the raw scores.

Expected logic:

- `home_score > away_score` means `match_result == "H"` and `result_class == 1`
- `home_score == away_score` means `match_result == "D"` and `result_class == 0`
- `home_score < away_score` means `match_result == "A"` and `result_class == -1`
- `goal_difference` should equal `home_score - away_score`
- `total_goals` should equal `home_score + away_score`

```{r outcome-logic-check}
outcome_logic_check <- purrr::imap_dfr(
  match_tables,
  function(dat, dataset_name) {
    dat |>
      dplyr::mutate(
        expected_match_result = dplyr::case_when(
          home_score > away_score ~ "H",
          home_score == away_score ~ "D",
          home_score < away_score ~ "A",
          TRUE ~ NA_character_
        ),
        expected_result_class = dplyr::case_when(
          expected_match_result == "H" ~ 1L,
          expected_match_result == "D" ~ 0L,
          expected_match_result == "A" ~ -1L,
          TRUE ~ NA_integer_
        ),
        expected_home_win = as.integer(expected_match_result == "H"),
        expected_draw = as.integer(expected_match_result == "D"),
        expected_away_win = as.integer(expected_match_result == "A"),
        expected_goal_difference = home_score - away_score,
        expected_total_goals = home_score + away_score,
        result_logic_ok =
          match_result == expected_match_result &
          result_class == expected_result_class &
          home_win == expected_home_win &
          draw == expected_draw &
          away_win == expected_away_win &
          goal_difference == expected_goal_difference &
          total_goals == expected_total_goals
      ) |>
      dplyr::summarise(
        dataset = dataset_name,
        rows = dplyr::n(),
        failed_rows = sum(!result_logic_ok, na.rm = TRUE),
        missing_logic_rows = sum(is.na(result_logic_ok))
      )
  }
)

readr::write_csv(
  outcome_logic_check,
  file.path(NOTEBOOK_CSV_DIR, "08_outcome_logic_check.csv")
)

outcome_logic_check
```

```{r failed-outcome-examples}
failed_outcome_examples <- purrr::imap_dfr(
  match_tables,
  function(dat, dataset_name) {
    dat |>
      dplyr::mutate(
        expected_match_result = dplyr::case_when(
          home_score > away_score ~ "H",
          home_score == away_score ~ "D",
          home_score < away_score ~ "A",
          TRUE ~ NA_character_
        ),
        expected_result_class = dplyr::case_when(
          expected_match_result == "H" ~ 1L,
          expected_match_result == "D" ~ 0L,
          expected_match_result == "A" ~ -1L,
          TRUE ~ NA_integer_
        ),
        expected_home_win = as.integer(expected_match_result == "H"),
        expected_draw = as.integer(expected_match_result == "D"),
        expected_away_win = as.integer(expected_match_result == "A"),
        expected_goal_difference = home_score - away_score,
        expected_total_goals = home_score + away_score,
        result_logic_ok =
          match_result == expected_match_result &
          result_class == expected_result_class &
          home_win == expected_home_win &
          draw == expected_draw &
          away_win == expected_away_win &
          goal_difference == expected_goal_difference &
          total_goals == expected_total_goals
      ) |>
      dplyr::filter(!result_logic_ok | is.na(result_logic_ok)) |>
      dplyr::select(
        source_match_id,
        date,
        competition,
        home_team,
        away_team,
        home_score,
        away_score,
        match_result,
        expected_match_result,
        result_class,
        expected_result_class,
        home_win,
        expected_home_win,
        draw,
        expected_draw,
        away_win,
        expected_away_win,
        goal_difference,
        expected_goal_difference,
        total_goals,
        expected_total_goals
      ) |>
      dplyr::slice_head(n = 20) |>
      dplyr::mutate(dataset = dataset_name, .before = 1)
  }
)

readr::write_csv(
  failed_outcome_examples,
  file.path(NOTEBOOK_CSV_DIR, "08_failed_outcome_examples.csv")
)

failed_outcome_examples
```

# 9. Missingness summary

```{r missingness-summary}
missingness_summary <- purrr::imap_dfr(
  processed_data,
  function(dat, dataset_name) {
    tibble::tibble(
      dataset = dataset_name,
      column = names(dat),
      n_rows = nrow(dat),
      n_missing = purrr::map_int(dat, ~ sum(is.na(.x))),
      pct_missing = round(100 * n_missing / n_rows, 2)
    )
  }
) |>
  dplyr::arrange(dataset, dplyr::desc(pct_missing), column)

readr::write_csv(
  missingness_summary,
  file.path(NOTEBOOK_CSV_DIR, "09_missingness_summary.csv")
)

missingness_summary
```

# 10. Duplicate checks

Some repeated fixtures may be legitimate, but they should be inspected.

```{r duplicate-checks}
duplicate_checks <- purrr::imap_dfr(
  match_tables,
  function(dat, dataset_name) {
    duplicate_source_ids <- dat |>
      dplyr::count(source_match_id, name = "n") |>
      dplyr::filter(n > 1L)

    duplicate_fixture_keys <- dat |>
      dplyr::count(
        date,
        home_team,
        away_team,
        competition,
        name = "n"
      ) |>
      dplyr::filter(n > 1L)

    exact_duplicate_rows <- dat |>
      dplyr::count(
        dplyr::across(dplyr::everything()),
        name = "n"
      ) |>
      dplyr::filter(n > 1L)

    tibble::tibble(
      dataset = dataset_name,
      duplicate_source_match_ids = nrow(duplicate_source_ids),
      duplicate_date_home_away_competition_keys = nrow(duplicate_fixture_keys),
      exact_duplicate_rows = nrow(exact_duplicate_rows)
    )
  }
)

readr::write_csv(
  duplicate_checks,
  file.path(NOTEBOOK_CSV_DIR, "10_duplicate_checks.csv")
)

duplicate_checks
```

# 11. football-data.co.uk numeric stat validation

Some football-data.co.uk columns are read as doubles because historical files can be inconsistent. This check verifies whether those columns are still integer-valued.

```{r football-data-integer-valued-check}
football_data <- processed_data$football_data_uk_matches

integer_valued_stat_cols <- c(
  "half_time_home_score",
  "half_time_away_score",
  "home_shots",
  "away_shots",
  "home_shots_on_target",
  "away_shots_on_target",
  "home_corners",
  "away_corners",
  "home_fouls",
  "away_fouls",
  "home_yellow_cards",
  "away_yellow_cards",
  "home_red_cards",
  "away_red_cards"
)

integer_valued_stat_cols <- intersect(integer_valued_stat_cols, names(football_data))

integer_valued_check <- tibble::tibble(
  column = integer_valued_stat_cols,
  class = purrr::map_chr(
    football_data[integer_valued_stat_cols],
    ~ paste(class(.x), collapse = " / ")
  ),
  n_missing = purrr::map_int(
    football_data[integer_valued_stat_cols],
    ~ sum(is.na(.x))
  ),
  non_integer_values = purrr::map_int(
    football_data[integer_valued_stat_cols],
    ~ sum(!is.na(.x) & .x != floor(.x))
  ),
  min_value = purrr::map_dbl(
    football_data[integer_valued_stat_cols],
    ~ suppressWarnings(min(.x, na.rm = TRUE))
  ),
  max_value = purrr::map_dbl(
    football_data[integer_valued_stat_cols],
    ~ suppressWarnings(max(.x, na.rm = TRUE))
  )
)

readr::write_csv(
  integer_valued_check,
  file.path(NOTEBOOK_CSV_DIR, "11_football_data_integer_valued_check.csv")
)

integer_valued_check
```

# 12. Result distribution

```{r result-distribution}
result_distribution <- purrr::imap_dfr(
  match_tables,
  function(dat, dataset_name) {
    dat |>
      dplyr::count(match_result, name = "matches") |>
      dplyr::mutate(
        dataset = dataset_name,
        pct = round(100 * matches / sum(matches), 2),
        .before = 1
      )
  }
)

readr::write_csv(
  result_distribution,
  file.path(NOTEBOOK_CSV_DIR, "12_result_distribution.csv")
)

result_distribution
```

# 13. Score and goals summary

```{r score-summary}
score_summary <- purrr::imap_dfr(
  match_tables,
  function(dat, dataset_name) {
    dat |>
      dplyr::summarise(
        dataset = dataset_name,
        rows = dplyr::n(),
        min_home_score = min(home_score, na.rm = TRUE),
        max_home_score = max(home_score, na.rm = TRUE),
        min_away_score = min(away_score, na.rm = TRUE),
        max_away_score = max(away_score, na.rm = TRUE),
        mean_home_score = round(mean(home_score, na.rm = TRUE), 3),
        mean_away_score = round(mean(away_score, na.rm = TRUE), 3),
        mean_total_goals = round(mean(total_goals, na.rm = TRUE), 3),
        max_total_goals = max(total_goals, na.rm = TRUE)
      )
  }
)

readr::write_csv(
  score_summary,
  file.path(NOTEBOOK_CSV_DIR, "13_score_summary.csv")
)

score_summary
```

# 14. Extreme score examples

Extreme rows are useful for catching parsing problems or unusual but real matches.

```{r extreme-score-examples}
extreme_score_examples <- purrr::imap_dfr(
  match_tables,
  function(dat, dataset_name) {
    dat |>
      dplyr::arrange(dplyr::desc(total_goals)) |>
      dplyr::select(
        source_match_id,
        date,
        competition,
        home_team,
        away_team,
        home_score,
        away_score,
        match_result,
        result_class,
        total_goals,
        goal_difference,
        neutral
      ) |>
      dplyr::slice_head(n = 10) |>
      dplyr::mutate(dataset = dataset_name, .before = 1)
  }
)

readr::write_csv(
  extreme_score_examples,
  file.path(NOTEBOOK_CSV_DIR, "14_extreme_score_examples.csv")
)

extreme_score_examples
```

# 15. Column display audit

This section shows which columns were omitted from compact example tables. They are still loaded in `processed_data`; they were only omitted from compact displays.

```{r displayed-versus-omitted-columns}
displayed_columns <- list(
  statsbomb_competitions = c(
    "competition_id",
    "season_id",
    "country_name",
    "competition_name",
    "competition_gender",
    "competition_youth",
    "competition_international",
    "season_name"
  ),

  statsbomb_matches = c(
    "source_match_id",
    "date",
    "season",
    "competition",
    "home_team",
    "away_team",
    "home_score",
    "away_score",
    "match_result",
    "result_class",
    "neutral"
  ),

  football_data_uk_matches = c(
    "source_match_id",
    "date",
    "season",
    "competition",
    "home_team",
    "away_team",
    "home_score",
    "away_score",
    "match_result",
    "result_class",
    "full_time_result",
    "home_shots",
    "away_shots",
    "home_shots_on_target",
    "away_shots_on_target"
  ),

  international_results = c(
    "source_match_id",
    "date",
    "season",
    "tournament",
    "home_team",
    "away_team",
    "home_score",
    "away_score",
    "match_result",
    "result_class",
    "neutral",
    "city",
    "country"
  )
)

omitted_columns_audit <- purrr::imap_dfr(
  processed_data,
  function(dat, dataset_name) {
    shown <- displayed_columns[[dataset_name]]
    all_cols <- names(dat)

    tibble::tibble(
      dataset = dataset_name,
      n_total_columns = length(all_cols),
      n_displayed_columns = length(intersect(shown, all_cols)),
      n_omitted_columns = length(setdiff(all_cols, shown)),
      omitted_columns = paste(setdiff(all_cols, shown), collapse = ", ")
    )
  }
)

readr::write_csv(
  omitted_columns_audit,
  file.path(NOTEBOOK_CSV_DIR, "15_omitted_columns_audit.csv")
)

omitted_columns_audit
```

# 16. Final checklist

After knitting this notebook, check:

1. All expected processed files exist.
2. All datasets load successfully.
3. Dates are parsed as `Date`.
4. Team, competition, tournament, city, and country fields are character columns.
5. Scores and result flags are integer fields where expected.
6. `neutral` is logical.
7. `match_result`, `result_class`, `home_win`, `draw`, and `away_win` agree with the scores.
8. `goal_difference` equals `home_score - away_score`.
9. `total_goals` equals `home_score + away_score`.
10. Missingness is explainable.
11. Duplicate checks do not reveal unexpected duplicate match identities.
12. Random and extreme examples look like real soccer match records.

# 17. Saved output files

```{r saved-output-files}
saved_output_files <- tibble::tibble(
  file = fs::dir_ls(NOTEBOOK_OUTPUT_DIR, recurse = TRUE, type = "file")
)

readr::write_csv(
  saved_output_files,
  file.path(NOTEBOOK_CSV_DIR, "17_saved_output_files.csv")
)

saved_output_files
