---
title: "Notebook 2: International Modeling Table EDA"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 9,
  fig.height = 5
)

PROJECT_ROOT <- normalizePath(
  "~/Documents/Projects/worldcup-forecast-r",
  winslash = "/",
  mustWork = TRUE
)

setwd(PROJECT_ROOT)

source("src/00_project_setup.R")
source("src/01_packages.R")
source("src/02_helpers.R")

OUTPUT_DIR <- file.path(
  PROJECT_ROOT,
  "notebooks",
  "notebook_2_outputs"
)

CHECK_DIR <- file.path(
  OUTPUT_DIR,
  "modeling_table_checks"
)

dir.create(OUTPUT_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(CHECK_DIR, recursive = TRUE, showWarnings = FALSE)
```

# 1. Notebook purpose

This notebook inspects the first international modeling table:

```text
data/processed/international_modeling_table.csv
```

The goal is to verify that:

* the table loads correctly;
* the train/test split is correct;
* pre-match Elo ratings do not leak same-day or post-match information;
* rating coverage is good enough for a first model;
* the outcome distribution is sensible;
* a first-model candidate dataset can be created.

This notebook does **not** fit a model.

# 2. Load modeling table

```{r load-modeling-table}
modeling_path <- file.path(PROCESSED_DIR, "international_modeling_table.csv")

if (!file.exists(modeling_path)) {
  stop("Modeling file not found: ", modeling_path)
}

matches <- read_processed_csv(modeling_path) |>
  dplyr::mutate(
    date = as.Date(date),
    home_rating_date = as.Date(home_rating_date),
    away_rating_date = as.Date(away_rating_date),
    year = lubridate::year(date),
    decade = floor(year / 10) * 10,
    data_split = as.character(data_split),
    match_result = as.character(match_result),
    result_class = as.integer(result_class),
    neutral = as.logical(neutral),
    shootout_played = as.logical(shootout_played),
    home_rating_pre_match = as.numeric(home_rating_pre_match),
    away_rating_pre_match = as.numeric(away_rating_pre_match),
    rating_diff = as.numeric(rating_diff),
    rating_age_days_home = as.numeric(rating_age_days_home),
    rating_age_days_away = as.numeric(rating_age_days_away),
    rating_available = !is.na(home_rating_pre_match) & !is.na(away_rating_pre_match)
  )

dplyr::glimpse(matches)
```

# 3. Basic inventory

```{r basic-inventory}
basic_inventory <- tibble::tibble(
  metric = c(
    "rows",
    "columns",
    "min_date",
    "max_date",
    "unique_source_match_ids",
    "duplicate_source_match_ids",
    "unique_home_teams",
    "unique_away_teams",
    "unique_tournaments",
    "rows_with_both_ratings",
    "rating_coverage"
  ),
  value = c(
    nrow(matches),
    ncol(matches),
    as.character(min(matches$date, na.rm = TRUE)),
    as.character(max(matches$date, na.rm = TRUE)),
    dplyr::n_distinct(matches$source_match_id),
    sum(duplicated(matches$source_match_id)),
    dplyr::n_distinct(matches$home_team),
    dplyr::n_distinct(matches$away_team),
    dplyr::n_distinct(matches$tournament),
    sum(matches$rating_available),
    round(mean(matches$rating_available), 4)
  )
)

basic_inventory

readr::write_csv(
  basic_inventory,
  file.path(CHECK_DIR, "01_basic_inventory.csv")
)
```

# 4. Core validation checks

```{r core-validation-checks}
core_checks <- tibble::tibble(
  check = c(
    "duplicate_source_match_id",
    "missing_date",
    "missing_home_team",
    "missing_away_team",
    "invalid_result_class",
    "invalid_data_split"
  ),
  rows_affected = c(
    sum(duplicated(matches$source_match_id)),
    sum(is.na(matches$date)),
    sum(is.na(matches$home_team) | matches$home_team == ""),
    sum(is.na(matches$away_team) | matches$away_team == ""),
    sum(!matches$result_class %in% c(-1L, 0L, 1L), na.rm = TRUE),
    sum(!matches$data_split %in% c("train", "test"), na.rm = TRUE)
  )
)

core_checks

readr::write_csv(
  core_checks,
  file.path(CHECK_DIR, "02_core_validation_checks.csv")
)

if (any(core_checks$rows_affected > 0)) {
  stop("Core modeling-table validation failed. Inspect 02_core_validation_checks.csv.")
}
```

# 5. Train/test split check

```{r split-summary}
split_summary <- matches |>
  dplyr::count(data_split, name = "matches") |>
  dplyr::mutate(prop = matches / sum(matches))

split_summary

readr::write_csv(
  split_summary,
  file.path(CHECK_DIR, "03_split_summary.csv")
)
```

```{r split-date-check}
split_date_check <- matches |>
  dplyr::summarise(
    train_rows_on_or_after_2018 = sum(
      data_split == "train" & date >= as.Date("2018-01-01"),
      na.rm = TRUE
    ),
    test_rows_before_2018 = sum(
      data_split == "test" & date < as.Date("2018-01-01"),
      na.rm = TRUE
    ),
    train_min_date = min(date[data_split == "train"], na.rm = TRUE),
    train_max_date = max(date[data_split == "train"], na.rm = TRUE),
    test_min_date = min(date[data_split == "test"], na.rm = TRUE),
    test_max_date = max(date[data_split == "test"], na.rm = TRUE)
  )

split_date_check

readr::write_csv(
  split_date_check,
  file.path(CHECK_DIR, "04_split_date_check.csv")
)

if (
  split_date_check$train_rows_on_or_after_2018 > 0 ||
  split_date_check$test_rows_before_2018 > 0
) {
  stop("Train/test split date check failed.")
}
```

# 6. Pre-match Elo timing check

This is the most important leakage check. A valid pre-match rating must be dated strictly before the match date.

```{r elo-timing-check}
elo_timing_check <- matches |>
  dplyr::summarise(
    rows = dplyr::n(),
    home_same_day = sum(!is.na(home_rating_date) & home_rating_date == date),
    away_same_day = sum(!is.na(away_rating_date) & away_rating_date == date),
    home_after_match = sum(!is.na(home_rating_date) & home_rating_date > date),
    away_after_match = sum(!is.na(away_rating_date) & away_rating_date > date),
    home_age_nonpositive = sum(
      !is.na(rating_age_days_home) & rating_age_days_home <= 0
    ),
    away_age_nonpositive = sum(
      !is.na(rating_age_days_away) & rating_age_days_away <= 0
    ),
    min_home_age = min(rating_age_days_home, na.rm = TRUE),
    median_home_age = median(rating_age_days_home, na.rm = TRUE),
    max_home_age = max(rating_age_days_home, na.rm = TRUE),
    min_away_age = min(rating_age_days_away, na.rm = TRUE),
    median_away_age = median(rating_age_days_away, na.rm = TRUE),
    max_away_age = max(rating_age_days_away, na.rm = TRUE)
  )

elo_timing_check

readr::write_csv(
  elo_timing_check,
  file.path(CHECK_DIR, "05_elo_timing_check.csv")
)

if (
  elo_timing_check$home_same_day > 0 ||
  elo_timing_check$away_same_day > 0 ||
  elo_timing_check$home_after_match > 0 ||
  elo_timing_check$away_after_match > 0 ||
  elo_timing_check$home_age_nonpositive > 0 ||
  elo_timing_check$away_age_nonpositive > 0
) {
  stop("Elo timing check failed: same-day, after-match, or non-positive rating ages found.")
}
```

# 7. Rating-difference arithmetic check

```{r rating-arithmetic-check}
rating_arithmetic_check <- matches |>
  dplyr::filter(
    !is.na(home_rating_pre_match),
    !is.na(away_rating_pre_match),
    !is.na(rating_diff)
  ) |>
  dplyr::summarise(
    checked_rows = dplyr::n(),
    bad_rating_diff_rows = sum(
      abs(rating_diff - (home_rating_pre_match - away_rating_pre_match)) > 1e-9
    )
  )

rating_arithmetic_check

readr::write_csv(
  rating_arithmetic_check,
  file.path(CHECK_DIR, "06_rating_arithmetic_check.csv")
)

if (rating_arithmetic_check$bad_rating_diff_rows > 0) {
  stop("Rating arithmetic check failed.")
}
```

# 8. Rating coverage

```{r rating-coverage-summary}
rating_coverage_summary <- matches |>
  dplyr::summarise(
    rows = dplyr::n(),
    rows_with_home_rating = sum(!is.na(home_rating_pre_match)),
    rows_with_away_rating = sum(!is.na(away_rating_pre_match)),
    rows_with_both_ratings = sum(rating_available),
    missing_either_rating = sum(!rating_available),
    rating_coverage = mean(rating_available)
  )

rating_coverage_summary

readr::write_csv(
  rating_coverage_summary,
  file.path(CHECK_DIR, "07_rating_coverage_summary.csv")
)
```

```{r rating-coverage-by-split}
rating_coverage_by_split <- matches |>
  dplyr::group_by(data_split) |>
  dplyr::summarise(
    rows = dplyr::n(),
    rows_with_both_ratings = sum(rating_available),
    missing_either_rating = sum(!rating_available),
    rating_coverage = mean(rating_available),
    .groups = "drop"
  )

rating_coverage_by_split

readr::write_csv(
  rating_coverage_by_split,
  file.path(CHECK_DIR, "08_rating_coverage_by_split.csv")
)
```

```{r rating-coverage-by-decade}
rating_coverage_by_decade <- matches |>
  dplyr::group_by(decade) |>
  dplyr::summarise(
    rows = dplyr::n(),
    rows_with_both_ratings = sum(rating_available),
    missing_either_rating = sum(!rating_available),
    rating_coverage = mean(rating_available),
    .groups = "drop"
  ) |>
  dplyr::arrange(decade)

rating_coverage_by_decade

readr::write_csv(
  rating_coverage_by_decade,
  file.path(CHECK_DIR, "09_rating_coverage_by_decade.csv")
)
```

# 9. Outcome distribution

```{r outcome-by-split}
outcome_by_split <- matches |>
  dplyr::count(data_split, match_result, name = "rows") |>
  dplyr::group_by(data_split) |>
  dplyr::mutate(prop = rows / sum(rows)) |>
  dplyr::ungroup()

outcome_by_split

readr::write_csv(
  outcome_by_split,
  file.path(CHECK_DIR, "10_outcome_by_split.csv")
)
```

```{r outcome-plot}
p_outcome <- outcome_by_split |>
  ggplot2::ggplot(
    ggplot2::aes(
      x = match_result,
      y = prop
    )
  ) +
  ggplot2::geom_col() +
  ggplot2::facet_wrap(~ data_split) +
  ggplot2::scale_y_continuous(labels = scales::percent_format()) +
  ggplot2::labs(
    title = "Outcome distribution by split",
    x = "Match result",
    y = "Proportion"
  )

p_outcome

ggplot2::ggsave(
  filename = file.path(CHECK_DIR, "plot_01_outcome_distribution_by_split.png"),
  plot = p_outcome,
  width = 9,
  height = 5
)
```

# 10. Elo rating difference distribution

```{r rating-diff-summary}
rating_diff_summary <- matches |>
  dplyr::filter(rating_available) |>
  dplyr::summarise(
    rows = dplyr::n(),
    min_rating_diff = min(rating_diff, na.rm = TRUE),
    q25_rating_diff = quantile(rating_diff, 0.25, na.rm = TRUE),
    median_rating_diff = median(rating_diff, na.rm = TRUE),
    mean_rating_diff = mean(rating_diff, na.rm = TRUE),
    q75_rating_diff = quantile(rating_diff, 0.75, na.rm = TRUE),
    max_rating_diff = max(rating_diff, na.rm = TRUE)
  )

rating_diff_summary

readr::write_csv(
  rating_diff_summary,
  file.path(CHECK_DIR, "11_rating_diff_summary.csv")
)
```

```{r rating-diff-plot}
p_rating_diff <- matches |>
  dplyr::filter(
    data_split == "train",
    rating_available
  ) |>
  ggplot2::ggplot(
    ggplot2::aes(x = rating_diff)
  ) +
  ggplot2::geom_histogram(bins = 60) +
  ggplot2::labs(
    title = "Training-set Elo rating difference distribution",
    subtitle = "rating_diff = home pre-match Elo - away pre-match Elo",
    x = "Rating difference",
    y = "Matches"
  )

p_rating_diff

ggplot2::ggsave(
  filename = file.path(CHECK_DIR, "plot_02_rating_diff_train.png"),
  plot = p_rating_diff,
  width = 9,
  height = 5
)
```

# 11. Outcomes by Elo rating-difference bin

```{r rating-bin-outcomes}
rating_bin_outcomes <- matches |>
  dplyr::filter(
    data_split == "train",
    rating_available,
    !is.na(rating_diff),
    match_result %in% c("H", "D", "A")
  ) |>
  dplyr::mutate(
    rating_diff_bin = dplyr::case_when(
      rating_diff <= -300 ~ "<= -300",
      rating_diff <= -150 ~ "-300 to -151",
      rating_diff <= -50 ~ "-150 to -51",
      rating_diff < 50 ~ "-50 to 49",
      rating_diff < 150 ~ "50 to 149",
      rating_diff < 300 ~ "150 to 299",
      TRUE ~ ">= 300"
    ),
    rating_diff_bin = factor(
      rating_diff_bin,
      levels = c(
        "<= -300",
        "-300 to -151",
        "-150 to -51",
        "-50 to 49",
        "50 to 149",
        "150 to 299",
        ">= 300"
      )
    )
  ) |>
  dplyr::count(rating_diff_bin, match_result, name = "rows") |>
  dplyr::group_by(rating_diff_bin) |>
  dplyr::mutate(prop = rows / sum(rows)) |>
  dplyr::ungroup()

rating_bin_outcomes

readr::write_csv(
  rating_bin_outcomes,
  file.path(CHECK_DIR, "12_rating_bin_outcomes_train.csv")
)
```

```{r rating-bin-plot}
p_rating_bin <- rating_bin_outcomes |>
  ggplot2::ggplot(
    ggplot2::aes(
      x = rating_diff_bin,
      y = prop,
      fill = match_result
    )
  ) +
  ggplot2::geom_col(position = "fill") +
  ggplot2::scale_y_continuous(labels = scales::percent_format()) +
  ggplot2::labs(
    title = "Training-set outcomes by Elo rating-difference bin",
    x = "Home Elo - Away Elo",
    y = "Outcome share",
    fill = "Result"
  ) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 30, hjust = 1)
  )

p_rating_bin

ggplot2::ggsave(
  filename = file.path(CHECK_DIR, "plot_03_outcomes_by_rating_diff_bin_train.png"),
  plot = p_rating_bin,
  width = 10,
  height = 5
)
```

# 12. Missing rating audit

```{r missing-rating-summary}
missing_ratings <- matches |>
  dplyr::filter(!rating_available) |>
  dplyr::mutate(
    missing_home_rating = is.na(home_rating_pre_match),
    missing_away_rating = is.na(away_rating_pre_match)
  )

missing_rating_summary <- missing_ratings |>
  dplyr::summarise(
    rows = dplyr::n(),
    missing_home = sum(missing_home_rating),
    missing_away = sum(missing_away_rating),
    min_date = min(date, na.rm = TRUE),
    max_date = max(date, na.rm = TRUE)
  )

missing_rating_summary

readr::write_csv(
  missing_rating_summary,
  file.path(CHECK_DIR, "13_missing_rating_summary.csv")
)
```

```{r unmatched-team-counts}
unmatched_team_counts <- missing_ratings |>
  dplyr::select(
    home_team,
    away_team,
    home_rating_pre_match,
    away_rating_pre_match
  ) |>
  dplyr::mutate(row_id = dplyr::row_number()) |>
  tidyr::pivot_longer(
    cols = c(home_team, away_team),
    names_to = "side",
    values_to = "team"
  ) |>
  dplyr::mutate(
    missing_rating = dplyr::case_when(
      side == "home_team" ~ is.na(home_rating_pre_match),
      side == "away_team" ~ is.na(away_rating_pre_match),
      TRUE ~ FALSE
    )
  ) |>
  dplyr::filter(missing_rating) |>
  dplyr::count(team, sort = TRUE, name = "missing_rating_appearances")

head(unmatched_team_counts, 50)

readr::write_csv(
  unmatched_team_counts,
  file.path(CHECK_DIR, "14_unmatched_team_counts.csv")
)
```

# 13. Tournament flags

```{r tournament-flag-summary}
tournament_flag_summary <- matches |>
  dplyr::summarise(
    rows = dplyr::n(),
    friendlies = sum(is_friendly, na.rm = TRUE),
    world_cup = sum(is_world_cup, na.rm = TRUE),
    world_cup_qualifier = sum(is_world_cup_qualifier, na.rm = TRUE),
    continental_tournament = sum(is_continental_tournament, na.rm = TRUE),
    continental_qualifier = sum(is_continental_qualifier, na.rm = TRUE)
  )

tournament_flag_summary

readr::write_csv(
  tournament_flag_summary,
  file.path(CHECK_DIR, "15_tournament_flag_summary.csv")
)
```

```{r top-tournaments}
top_tournaments <- matches |>
  dplyr::count(tournament, sort = TRUE, name = "matches") |>
  dplyr::slice_head(n = 25)

top_tournaments

readr::write_csv(
  top_tournaments,
  file.path(CHECK_DIR, "16_top_tournaments.csv")
)
```

# 14. Raw Elo file inspection

This confirms the raw Elo file schema. It is useful because the model table should use only lagged, pre-match ratings.

```{r raw-elo-inspection}
ratings_raw_path <- file.path(
  RAW_DIR,
  "international_ratings",
  "world_football_elo.csv"
)

if (!file.exists(ratings_raw_path)) {
  stop("Raw Elo file not found: ", ratings_raw_path)
}

ratings_raw <- readr::read_csv(
  ratings_raw_path,
  show_col_types = FALSE
)

names(ratings_raw)

dplyr::glimpse(ratings_raw)

head(ratings_raw, 20)
```

# 15. Candidate first-model dataset

The first candidate model dataset uses only rows where both teams have pre-match Elo ratings and the result is one of home win, draw, or away win.

```{r candidate-first-model-dataset}
first_model_data <- matches |>
  dplyr::filter(
    rating_available,
    match_result %in% c("H", "D", "A"),
    !is.na(rating_diff),
    data_split %in% c("train", "test")
  )

first_model_summary <- first_model_data |>
  dplyr::count(data_split, name = "rows") |>
  dplyr::mutate(prop = rows / sum(rows))

first_model_summary

readr::write_csv(
  first_model_summary,
  file.path(CHECK_DIR, "17_first_model_summary.csv")
)

readr::write_csv(
  first_model_data,
  file.path(OUTPUT_DIR, "international_first_model_candidate.csv")
)
```

# 16. Final pass/fail message

```{r final-message}
cat("\nMODELING TABLE CHECK PASSED\n")
cat("Rows:", nrow(matches), "\n")
cat("Rows with both ratings:", sum(matches$rating_available), "\n")
cat("Rating coverage:", round(mean(matches$rating_available) * 100, 2), "%\n")
cat("First-model rows:", nrow(first_model_data), "\n")
cat("First-model train rows:", sum(first_model_data$data_split == "train"), "\n")
cat("First-model test rows:", sum(first_model_data$data_split == "test"), "\n")
cat("Outputs written to:", OUTPUT_DIR, "\n")
cat("Check files written to:", CHECK_DIR, "\n")
```

# 17. Modeling notes

```{r write-notes}
eda_notes <- c(
  "# Notebook 2 EDA Notes",
  "",
  "This notebook inspects `data/processed/international_modeling_table.csv`.",
  "",
  "Main validation findings to confirm:",
  "- `source_match_id` is unique.",
  "- `data_split` follows the chronological split.",
  "- Elo ratings are strictly pre-match.",
  "- No same-day or post-match Elo ratings are used.",
  "- `rating_diff` equals home pre-match Elo minus away pre-match Elo.",
  "",
  "Main modeling-safe rule:",
  "- Use the train split for model development.",
  "- Keep the 2018+ test split untouched for final evaluation.",
  "",
  "First candidate model dataset:",
  "- Rows with both home and away pre-match Elo ratings.",
  "- Valid `match_result` in H/D/A.",
  "- Non-missing `rating_diff`.",
  "",
  "Recommended first model:",
  "- Multinomial logistic regression.",
  "- Target: `match_result`.",
  "- Predictors: `rating_diff`, `neutral`, and tournament flags.",
  "",
  "Do not use post-match columns as predictors:",
  "- `home_score`",
  "- `away_score`",
  "- `goal_difference`",
  "- `total_goals`",
  "- `shootout_winner`",
  "- `home_won_shootout`",
  "- `away_won_shootout`"
)

writeLines(
  eda_notes,
  con = file.path(OUTPUT_DIR, "notebook_2_eda_notes.md")
)

eda_notes
```

